---
title: "trans- to cis-eQTLs "
output: html_notebook
---

Are there strong trans-eQTLs when using Rnor_6.0 that turn out to be cis-eQTLs when using mRatBN7.2? Using the chromosome segments that remapped to another chromosome, look for cases where an eSNP occurs in a remapped segment but the eGene's TSS doesn't, or vice versa.

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(patchwork)

eqtls <- read_tsv("data/eqtl_moved.txt.gz", col_types = "ccdcll") |>
    mutate(moved = esnp_move | tss_move)

reloc <- read_tsv("data/relocated.txt", col_types = "ccdcicicllcicic")

genes <- read_tsv("data/genes.txt", col_types = "ccc---i-----")
gene_symbols <- genes |>
    select(gene_id, gene_name) |>
    deframe()
```

Substantial differences between cis- and trans-eQTLs, though note that the 3537 pairs with relocated TSS contain just two unique eGenes.

```{r}
with(eqtls, table(esnp_move, tss_move, rn6_type))
```

Interpretation: cis-eSNPs are ~100x less likely than trans-eSNPs to have been mislocated. trans-eGene TSS are as likely if not more likely than trans-eSNPs to be mislocated, but cis-eGene TSS are never mislocated.

To check this, see where mislocated eSNPs and TSS relocate, and which trans-eQTLs become cis-eQTLs, and vice versa.

```{r}
with(reloc, table(rn6_type, rn7_type))
```

```{r}
eqtls |>
    left_join(select(reloc, gene_id, variant_id, rn7_type),
              by = c("gene_id", "variant_id"),
              relationship = "one-to-one") |>
    mutate(rn7_type = if_else(moved, rn7_type, "unchanged"),
           rn7_type = fct_relevel(rn7_type, "unchanged")) |>
    with(table(rn6_type, rn7_type))
```

How many eGenes have trans->cis-eQTLs?

```{r}
reloc |>
    filter(rn6_type == "trans",
           rn7_type == "cis") |>
    count(gene_id)
```

```{r}
eqtls |>
    ggplot(aes(x = rn6_type, fill = moved)) +
    geom_bar()
```

```{r}
eqtls |>
    filter(rn6_type == "trans") |>
    left_join(select(reloc, variant_id, gene_id, rn7_type),
              by = c("variant_id", "gene_id"),
              relationship = "one-to-one") |>
    mutate(log10pval = -log10(pval),
           eSNPs = case_when(
               moved & rn6_type == "trans" & rn7_type == "cis" ~ "trans->cis",
               moved & rn6_type == "trans" & rn7_type == "trans" ~ "trans->trans",
               !moved ~ "unchanged"
           )) |>
    ggplot(aes(x = gene_id, y = log10pval, color = eSNPs)) +
    geom_point(size = 0.5) +
    scale_color_manual(values = c("red", "black", "gray")) +
    theme_minimal() +
    theme(axis.text.x = element_blank()) +
    xlab("eGenes (for all trans-eQTLs with p-value < 1e-8)") +
    ylab(expression(-log[10]*"p-value"))
```

Top (by p-value) genes with trans->cis-eQTLs:

```{r}
reloc |>
    filter(rn6_type == "trans",
           rn7_type == "cis") |>
    summarise(lowest_pval = min(pval),
              n_trans_to_cis = n(),
              .by = gene_id) |>
    arrange(lowest_pval) |>
    mutate(gene_symbol = gene_symbols[gene_id], .after = gene_id)
```

rn6 TSS distance for cis->trans-eQTLs:

```{r}
reloc |>
    filter(rn6_type == "cis") |>
    mutate(tss_distance = rn6_esnp_pos - rn6_tss_pos) |>
    ggplot(aes(x = tss_distance)) +
    geom_histogram(bins = 100)
```

rn7 TSS distance for trans->cis-eQTLs:

```{r}
reloc |>
    filter(rn7_type == "cis") |>
    mutate(tss_distance = rn7_esnp_pos - rn7_tss_pos) |>
    ggplot(aes(x = tss_distance)) +
    geom_histogram(bins = 100)
```
